\section{Constructional transparency}\label{sec:3}

Constructions vary in how much of their internal form--meaning organization remains accessible to speakers, learners, and analysts. This is the first of three sections on the form--meaning accessibility family: §\ref{sec:3} sets up the theoretical compositionality gradient and the central claim that constructional transparency is a network property; §\ref{sec:4} adds rated compound transparency; §\ref{sec:5} adds time-sensitive lexical processing. The three are within-domain profiles distinct in items and methods; whether they cohere into a cross-domain projectibility cluster is for §\ref{sec:7}.

\subsection{Constructions as form--meaning pairings}

Construction Grammar (henceforth CxG) treats constructions, learned form--meaning pairings, as the basic units of grammar \autocite{goldberg-2006-constructions-at-work}. Constructions exist at all levels: morphemes (\mention{pre-}, \mention{-ing}); words (\mention{avocado, anaconda, and}); complex words (\mention{daredevil, shoo-in}); the regular-plural pattern [N\mention{-s}]; filled idioms (\mention{going great guns, give the Devil his due}); partly-filled idioms (\mention{jog} \(\langle\textup{someone's}\rangle\) \mention{memory}, \mention{send} \(\langle\textup{someone}\rangle\) \mention{to the cleaners}); Covariational Conditional (\mention{The} X\mention{-er the} Y\mention{-er}, e.g., \mention{the more you think about it, the less you understand}); Ditransitive (Subj V Obj1 Obj2); Passive \autocite[5, Table 1.1]{goldberg-2006-constructions-at-work}. The list is heterogeneous; the unifying claim is that grammatical knowledge consists of these pairings, not of derivations applied to abstract underlying structures.

Argument structure isn't provided by verbs alone but by argument-structure constructions that combine with verbs. Goldberg's worked example uses \mention{slice} across five argument-structure constructions \autocite[6--7]{goldberg-2006-constructions-at-work}.

\ea\label{ex:slice-constructions}
\ea \mention{He sliced the bread.}\hfill[\textsc{transitive}]
\ex \mention{Pat sliced the carrots into the salad.}\hfill[\textsc{caused motion}]
\ex \mention{Pat sliced Chris a piece of pie.}\hfill[\textsc{ditransitive}]
\ex \mention{Emeril sliced and diced his way to stardom.}\hfill[\mention{way} \textsc{construction}]
\ex \mention{Pat sliced the box open.}\hfill[\textsc{resultative}]
\z
\z

\mention{Slice} contributes a constant lexical meaning across all five examples in (\ref{ex:slice-constructions}). The argument-structure semantics (someone acting on something; causing something to move; intending someone to receive something; moving despite obstacles; causing change of state) comes from the constructions.

Two further commitments follow. CxG is \term{usage-based}: patterns are stored as constructions if some aspect is non-predictable, or if they're fully predictable but sufficiently frequent \autocite[5]{goldberg-2006-constructions-at-work}. The framework is non-derivational: \enquote{what you see is what you get,} with no underlying levels, no movement, and no phonologically null elements \autocite[11]{goldberg-2006-constructions-at-work}. Both commitments matter for the transparency question, since they determine what counts as a construction's \enquote{parts} and what counts as the construction's contribution.

In the umbrella schema's vocabulary: $X$ (the mediator) is the construction; $P$ (what remains accessible) is the meanings of the parts and their compositional contribution to the whole; $R$ (the relation) is interpretation, learning, and analogical generalization. The relation is broader than §\ref{sec:2}'s agreement, and the accessibility is gradient rather than class-like.

\subsection{The compositionality gradient}\label{sec:3-2}

Compositionality is best treated not as a binary property but as a gradient. The working scale below discretizes that gradient into four points for expository convenience. CxG itself presents compositionality as continuous; the four-point scheme is borrowed from the broader idiom-classification tradition (which distinguishes fully compositional from partly schematic and idiomatic constructions), not from CxG's own architecture, and is used here as a presentation aid rather than a CxG-internal claim.

{\def\LTcaptype{none} % don't increment counter
\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\linewidth - 2\tabcolsep) * \real{0.3619}}
  >{\raggedright\arraybackslash}p{(\linewidth - 2\tabcolsep) * \real{0.6381}}@{}}
\toprule\noalign{}
\begin{minipage}[t]{\linewidth}\raggedright
Type
\end{minipage} & \begin{minipage}[t]{\linewidth}\raggedright
Transparency profile
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Fully compositional pattern & parts and arrangement predict the interpretation \\
Partly schematic construction & construction contributes meaning, but parts remain interpretable \\
Idiomatically constrained construction & construction-specific meaning partly overrides compositional access \\
Frozen idiom & parts have little current interpretive accessibility \\
\end{longtable}
}

At the \term{fully compositional} end, simple transitive sentences like the transitive \mention{slice} example above have parts whose meanings predict the whole. The construction contributes argument linking, tense, and agreement, but the propositional content reduces to verb-meaning plus arguments. Fully compositional cases are the limiting endpoint of form--meaning transparency, not the paradigm. The transitive construction-type admits a default-conceal variant (frozen idioms like \mention{kick the bucket} conceal compositional access), so any transitive sentence sits on a gradient where concealment is a real option for the construction-type; fully compositional sentences register as transparency at the trivial limit of that gradient. This is what differentiates them from §\ref{sec:2-3}'s coordination case: the coordinate construction-type admits no default-conceal variant, so coordination doesn't sit on the gradient at all and the schema excludes it.

\term{Partly schematic constructions} are where CxG does most of its empirical work. The Ditransitive (Subj V Obj1 Obj2) is the canonical example \autocite[10]{goldberg-2006-constructions-at-work}.

\begin{samepage}
\ea\label{ex:ditransitive-animacy}
\ea \mention{Liza bought Zach a book.}
\ex \mention{Liza sent Stan a book.}
\ex \marg{?\mention{Liza sent storage a book.}}
\z
\z
\end{samepage}

The first two examples in (\ref{ex:ditransitive-animacy}) both inherit the construction's meaning of intended transfer to an animate recipient, even though \mention{buy} and \mention{send} don't lexically encode that meaning. The animacy constraint on the recipient explains the contrast between the second and third examples. The construction contributes transfer-with-animate-recipient meaning; the verb contributes its lexical content. The verb and construction are interpretable independently and combine compositionally, but the combination is constructional, not just lexical.

\term{Idiomatically constrained constructions} include partly-filled idioms (\mention{jog} \(\langle\textup{someone's}\rangle\) \mention{memory}; \mention{send} \(\langle\textup{someone}\rangle\) \mention{to the cleaners}). The construction has a fixed lexical core whose meaning is metaphorical or non-literal, but the open slot is fillable by ordinary NPs and the surrounding syntax is regular. \mention{Memory} in \mention{jog memory} is interpretable as the same noun that appears elsewhere; \mention{jog} contributes a metaphor of brief shaking. The construction-specific meaning (`cause to recall briefly') partly overrides the compositional reading, but the parts remain visible. Schema-wise: the construction is the mediator; the parts' meanings remain partly accessible; the relation is interpretation, but the interpretation is partly construction-specific.

\term{Frozen idioms} (\mention{kick the bucket}) test the lower bound. The parts contribute nothing to the construction's meaning of `die'; \mention{kick} and \mention{bucket} are dead lexically inside the construction even if they remain alive elsewhere. At this end of the gradient, $P$ (the parts' meanings) is no longer accessible through $X$ (the construction) for $R$ (interpretation). The construction passes the schema's mediator-of-form requirement (it has internal structure that \mention{could} have mediated access) but fails the accessibility-remains-through-it test. Frozen idioms mark where transparency in the umbrella sense bottoms out for the interpretation relation.

This gradient isn't unique to argument-structure or idiomatic constructions; it runs through morphology too. Productive prefixation (\mention{un-} in \mention{unhappy}) sits near the fully compositional end. Partly opaque derivatives (\mention{business} from \mention{busy}) sit near the idiomatic end: the historical compositional source is recoverable for analysts but not for ordinary speakers. The same gradient logic recurs in §\ref{sec:4} (compound transparency) and §\ref{sec:5} (priming-graded morphological transparency), though the domains and operationalizations differ.

\subsection{Constructional transparency as a network property}\label{sec:3-3}

Section \ref{sec:3-2} treated each point on the gradient as a property of an individual construction. This paper develops a stronger claim that builds on Goldberg's framework without being orthodoxy in it: constructional transparency also depends on the network in which a construction sits. Inheritance hierarchies link broad generalizations to subregularities to exceptional low-level constructions \autocite[14]{goldberg-2006-constructions-at-work}. A construction's parts and their compositional contribution are accessible partly because related constructions in the network make them so.

Goldberg's worked example is the \mention{What's X doing Y?} construction \autocite[14]{goldberg-2006-constructions-at-work}, which she credits to Kay and Fillmore \autocite{kay-fillmore-1999-whats-x-doing-y}.

\ea\label{ex:whats-doing}
\mention{What's that fly doing in my soup?!}
\z

The construction in (\ref{ex:whats-doing}) has a fixed form and connotes unexpectedness; it inherits from the Left Isolation construction, the Subject-Auxiliary Inversion construction, the Subject-Predicate construction, and the Verb-Phrase construction. A learner who treats \mention{What's the} X \mention{doing} Y\mention{?!} as a construction in its own right does so partly because each of its inherited constructions has its own attestations and contributes a piece of the form--meaning organization. Without the inherited constructions in the network, the learner would have less to work with.

This makes constructional transparency a network property in two distinct senses. \term{Interpretive transparency}: a construction's compositional reading is accessible partly because related constructions provide compositional support. The Ditransitive's transfer meaning is interpretable partly because the \mention{for}-dative paraphrase and the regular argument-structure constructions provide compositional context \autocite[10]{goldberg-2006-constructions-at-work}.

\ea\label{ex:for-dative}
\mention{Liza bought a book for Zach.}
\z

The \mention{for}-dative in (\ref{ex:for-dative}) supplies the relevant paraphrase relation. \term{Learnability transparency}: a construction is acquirable partly because its place in the inheritance network gives the learner multiple routes of access. Both senses are $R$-relativized in the umbrella schema: $P$ remains accessible for the purposes of $R$, where $R$ is interpretation in the first sense and learning or analogical generalization in the second.

The network claim is offered as a hypothesis, not as established. If the hypothesis holds, constructions in dense regions of the constructicon would be more transparent for both $R$-values than constructions in sparse regions, and constructions whose family-resemblance relatives have low type frequency would be harder to interpret and learn. Goldberg \autocite*[15]{goldberg-2006-constructions-at-work} cites usage-based acquisition work (Diessel and Tomasello, Tomasello, Saffran and colleagues) as part of the broader case for usage-based acquisition; whether any of those studies bear specifically on transparency-as-network-property is itself an empirical question this paper leaves open. The form--meaning accessibility family of §§\ref{sec:3}--\ref{sec:5} will return to the hypothesis: the corpus-derivable predictors that Bell and Schäfer use in §\ref{sec:4} and the priming-magnitude modulations that Heyer and Kornishova report in §\ref{sec:5} will be treated as compatible quantitative probes for the network-relative transparency that §\ref{sec:3} frames theoretically, not as direct measurements of it.

The network claim also clarifies what frozen idioms mark. \mention{Kick the bucket} has nearly no parts-to-whole compositional accessibility, but speakers recognize it as an idiom rather than as random word combinations. That recognition is partly supported by the existence of partly-idiomatic constructions one rung up the gradient: network position helps explain why the string is treated as a unit. The construction can also be stored holistically and accessed without compositional decomposition; nothing in the network claim rules that out. Even at the bottom of the gradient, network position does some work, but the work is recognition-relative, not interpretation-relative.

\subsection{Number-transparent NPs revisited}\label{sec:3-4}

Section \ref{sec:2} treated \textit{CGEL} number-transparent NPs as agreement-accessibility cases. From the CxG side, those same NPs are constructions: form--meaning pairings whose form is the number-transparent quantificational noun + \mention{of}-complement template and whose constructional behaviour includes the agreement override. \textit{CGEL} preserves head-and-complement structure (with \mention{lot, number}, etc.~as heads) and treats the override as a property of the construction rather than a separately stipulated mechanism (the term \mention{percolation} appears only informally in a footnote, §18.2, fn.~12). CxG variants differ on whether head-and-complement structure is preserved internally: Sign-Based Construction Grammar (SBCG) preserves it explicitly; some Goldbergian variants leave the internal representation more abstract. The choice is methodological; the umbrella claim survives under any of them.

This connects §\ref{sec:2} and §\ref{sec:3} in a useful way. Number-transparent NPs are constructional-transparency cases, but the relation $R$ is agreement (a §\ref{sec:2}-style morphosyntactic relation), not interpretation (the §\ref{sec:3}-style relation). The umbrella schema is one frame; the senses partition by $R$. A construction can be transparent for one $R$ and opaque for another: a frozen idiom like \mention{kick the bucket} is opaque for interpretation but morphosyntactically regular (the verb agrees with its subject in the ordinary way). $R$-relativity is built into the umbrella; it's a feature, not a bug. The schema asks whether $P$ remains accessible \mention{for the purposes of R}, and accessibility isn't transitive across $R$-values.

The methodological consequence: choosing \textit{CGEL}'s apparatus or CxG's apparatus for any given case is a question about which formal vocabulary makes the empirical generalizations clean. The deeper question of which framework is \enquote{right} isn't at issue here. The umbrella schema sits above both: it states the relational claim that a transparency analysis has to satisfy, without committing to how the satisfaction is formally encoded.

A coherence point across the frameworks the paper engages: \textit{CGEL}, SBCG, and Goldbergian CxG are all non-derivational, and none posits empty mediating elements (traces, copies) between a surface position and a hypothesized underlying one. Contemporary dependency-grammar accounts \autocite{gibson2025,dacunha-gibson-2025-syntactic-complexity} take the same line on long-distance dependencies, arguing that direct head-to-dependent connections fit acceptability and processing data without empty mediators (cf.~the §\ref{sec:2} extraction-transparency footnote). The umbrella schema doesn't adjudicate the dispute, but the frameworks engaged here all treat the mediator $X$ as observable structure rather than as posited invisible mediation between observable layers. Where the schema asks whether $P$ remains accessible through $X$ for $R$, the candidate $X$ is always something the analyst can point at.

\subsection{Transparent free relatives revisited}\label{sec:3-5}

The TFR profile is constructional transparency in the sense needed here: the parent/daughter contrast is an access condition inherited through a constructional network, not a per-item lexical property. The parent licenses agreement transparency; the attributional daughter licenses syntactic-category transparency. That is the profile-level point §\ref{sec:3} needs. The fuller SBCG architecture, verb-class corpus check, PP-nucleus restriction, and competing metalinguistic versus mention-mediated readings are deferred to Kim's \autocite*[ch.~6.8]{kim-2026-form-function} treatment and Kim and Reynolds \autocite*{kim-reynolds-2026-tfr-two-kinds}.

\subsection{Looking ahead to §§\ref{sec:4}--\ref{sec:5}}

Section \ref{sec:3} has set up the theoretical frame for the form--meaning accessibility family: constructions as form--meaning pairings, compositionality as gradient, transparency as network property. Section 4 will give the quantitative measurement. Bell and Schäfer's \autocite*{bell-schafer-2016-modelling-transparency} models of compound transparency operationalize expectedness via corpus statistics and human ratings, providing empirical content for the partly-schematic and idiomatic regions of the gradient. Section 5 will give the processing measurement. Heyer and Kornishova \autocite*{heyer-kornishova-2018-priming-eventually} show that semantic transparency modulates masked priming, but only at long stimulus-onset asynchronies, a finding that bears on the time-course of network access.

The three sections together motivate the form--meaning accessibility family, with three evidential windows that bear on it; whether they converge into a cross-domain projectibility cluster is what §\ref{sec:7} will weigh, and §\ref{sec:7} doesn't claim convergence delivered. Whether each within-domain profile passes Slater's \autocite*{slater-2015-natural-kindness} SPC test, and whether Khalidi's \autocite*{khalidi-2018-causal-networks} causal-network refinement applies to any cross-domain reading, are questions §\ref{sec:7} returns to. For now, §\ref{sec:3} has done its job by setting the frame: constructional transparency is gradient and network-relative, and the next two sections give the family its quantitative and processing-experimental bone structure.
