\section{Processing transparency}\label{sec:5}

Section 4's rated transparency is a steady-state property of the compound. Section 5 turns to a time-course property: how does semantic transparency modulate the processing of morphologically complex words during recognition? Heyer and Kornishova's \autocite*{heyer-kornishova-2018-priming-eventually} masked-priming study is the empirical anchor. Their finding, summarized in their title's three-dot ellipsis: semantic transparency affects morphological priming, but only at long stimulus-onset asynchronies.

\subsection{The Heyer and Kornishova study}

Heyer and Kornishova \autocite*{heyer-kornishova-2018-priming-eventually} ran two parallel masked-priming experiments, one with English -\mention{ness} nominalizations and one with Russian -\mention{ost'} nominalizations. The two suffixes are deadjectival and produce abstract nouns denoting `state of being X' (e.g., \mention{dark} + \mention{ness} = `darkness'; \mention{bledn-} + \mention{-ost'} = \mention{blednost'} `paleness'). Targets in both experiments were the adjective bases (e.g., \mention{pale, dark}); primes were the corresponding nominalization, a simplex unrelated control, or the adjective as identity prime. Each experiment crossed two stimulus-onset asynchronies (SOAs), short and long, with transparency as a continuous variable rated by separate groups of speakers (Appendix~\ref{sec:C}).

Transparency was rated on a 1--7 scale: \mention{business} (rated 3.17) at the opaque end is no longer interpretable as `state of being busy'; \mention{falseness} (rated 5.75) is fully derivable from \mention{false}; items spanned the range continuously.

The headline result is SOA-conditioned. At short SOA in both languages, priming is constant across the transparency scale: morphologically complex primes are decomposed into stems regardless of whether the stem still contributes to the meaning of the whole. At long SOA, priming becomes transparency-modulated: more transparent prime-target pairs (e.g., \mention{paleness-pale}) yield larger priming effects than less transparent pairs (e.g., \mention{business-busy}). The three-way interaction (Prime Type × Transparency × SOA) is significant in English and marginally so in Russian, with the same direction in both languages \autocite[1117--1119]{heyer-kornishova-2018-priming-eventually}. Statistical detail is in Appendix~\ref{sec:C-3}.

\subsection{The serial-processing interpretation}

Heyer and Kornishova read the data as evidence for a serial decomposition process. Morphologically complex words are first decomposed into stem-plus-affix on the basis of orthographic structure (\mention{morpho-orthographic decomposition}), with semantic information entering only at a later stage. The first stage is semantically blind: it parses \mention{paleness} into \mention{pale} + \mention{ness} regardless of whether the parse yields a coherent semantic interpretation, and would parse \mention{corner} into \mention{corn} + \mention{er} even though the parse fails semantically \autocite[1112]{heyer-kornishova-2018-priming-eventually}.

This pattern is consistent with Bell and Schäfer's \autocite*{bell-schafer-2016-modelling-transparency} hypothesis that \enquote{perceived transparency may itself be a reflex of ease of processing.} Within the masked-priming paradigm, fast-and-shallow windows show no transparency modulation; longer windows do. The two studies probe different items (NN compounds, derivational nominalizations) and different methods (ratings, masked priming), so they don't test the same processing chain; they're compatible probes for a shared hypothesis.

\subsection{The rival parallel-processing position}

The serial reading is contested. Feldman and colleagues' parallel-immediate position argues that morpho-semantic information influences recognition from the earliest stages. Feldman et al.'s \autocite*{feldman-etal-2015-must-analysis} time-course analysis presses the diagnostic question (\mention{must analysis of meaning follow analysis of form?}) with a same-target design (each target is compared after semantically similar and semantically dissimilar primes within subjects) across SOAs of 34, 48, 67, 84, and 100~ms. They report reliable semantic effects at SOAs as short as 34~ms (their solo-SOA Experiment 3) and 48~ms (Experiment 2); the 34~ms lag is at or below Heyer and Kornishova's short windows (39~ms English, 33~ms Russian). The effect is small at the shortest lag in the intermixed-SOA design (a 6~ms similar-versus-dissimilar difference, with the slope interaction only marginal, $\chi^2(1) = 3.08$, $p = 0.08$), which is the room Heyer and Kornishova use to read the short-SOA evidence as too weak to overturn a serial account. On the parallel-immediate reading, the form-then-meaning ordering doesn't survive Feldman et al.'s short-SOA effects; on Heyer and Kornishova's, it does.

These two designs aren't head-to-head: Heyer and Kornishova's short-SOA result is the absence of transparency modulation across an item set rather than the absence of semantic effects on a same-target comparison, and the disagreement is whether semantic information is constitutively present at masked-priming windows or emerges only after morpho-orthographic decomposition. The literature is mixed enough (the meta-table on p.~1121 of Heyer and Kornishova surveys 35 prior masked-priming studies) that this paper takes the serial reading as Heyer and Kornishova's reading, not as a settled disciplinary verdict. The cross-modal probe in Appendix~\ref{sec:D} would generate within-items evidence relevant to the disagreement.

\subsection{What §\ref{sec:5} contributes to the cluster}

Section \ref{sec:5} contributes the time-course dimension to the form--meaning accessibility family. Section \ref{sec:3} framed compositional transparency as a gradient; Section \ref{sec:4} made it quantitative through corpus predictors and ratings; Section \ref{sec:5} makes it temporally articulated. The family's evidential support is now: a theoretical scale, a steady-state quantitative measure, and a time-course measure. The three windows are operationally distinct but conceptually related: they're different probes for the same underlying form--meaning accessibility, taken at different stages of speakers' interaction with the relevant items. Falsification condition for the within-domain priming profile: if rated semantic transparency failed to modulate masked-priming magnitude at long SOAs in an extension or replication of Heyer and Kornishova's design, the profile would fail.

Bell and Schäfer's compounds and Heyer and Kornishova's nominalizations are different items in different morphological domains; their results don't compose into a transitive empirical chain on the same items. They're compatible probes for the form--meaning accessibility hypothesis, not direct measurements of one another. Whether the cluster hypothesis survives the within-items test (corpus predictors, ratings, priming magnitude, and interpretation accuracy aligning on shared items as joint reflexes of underlying representational structure) is the empirical question Appendix~\ref{sec:D} sketches as a probe design. Until that probe is run, the family is three within-domain profiles plus a research hypothesis, not a delivered cross-domain cluster.

\subsection{Cluster status and bridge to §\ref{sec:6}}

Sections \ref{sec:3}--\ref{sec:5} give three within-domain projectibility profiles stabilized by different sources: constructional inheritance (§\ref{sec:3}), usage-based entrenchment (§\ref{sec:4}), processing architecture (§\ref{sec:5}). Whether they compose into a cross-domain cluster on shared items is for §\ref{sec:7}, with the probe design in Appendix~\ref{sec:D}. Acquisition trajectory is a plausible further extension consistent with §\ref{sec:3-3}'s network claim but not directly grounded by these sections. Section \ref{sec:6} turns from form--meaning accessibility to the comparative question of §\ref{sec:1}.
