\section{Compound transparency: Bell and Schäfer's empirical apparatus}

This appendix carries the detailed exposition of Bell and Schäfer \autocite*{bell-schafer-2016-modelling-transparency} that §\ref{sec:4} references in compressed form. The body keeps the headline finding (three predictors, scalar transparency, within-domain projection); the predicate inventory and stress speculation live here.

\subsection{Levi's predicate inventory}\label{sec:B-1}

Bell and Schäfer code the semantic relation between the constituents of each NN compound using Levi's (1978) nine predicates: CAUSE, HAVE, MAKE, USE, BE, IN, FOR, FROM, ABOUT \autocite[167]{bell-schafer-2016-modelling-transparency}.

Three of the predicates (CAUSE, HAVE, MAKE) come in two variants depending on which constituent functions as the subject of the predicate:

\begin{itemize}
\tightlist
\item
  CAUSE1 \mention{tear gas} (\enquote{gas that causes tears}) versus CAUSE2 \mention{drug deaths} (\enquote{deaths caused by drugs}).
\item
  HAVE1 \mention{picture book} versus HAVE2 \mention{government land}.
\item
  MAKE1 \mention{honeybee} versus MAKE2 \mention{daisy chains}.
\end{itemize}

The other six predicates (USE, BE, IN, FOR, FROM, ABOUT) don't subdivide.

\subsection{Metaphoric/metonymic shift coding}

Bell and Schäfer's earlier work coded each compound and each constituent for metaphoric and metonymic shifts, treating shift as a binary feature \autocite[167]{bell-schafer-2016-modelling-transparency}. The coding is independent of the predicate-relation assignment and contributes separately to predicting transparency ratings.

\subsection{IN, FOR, and the stress hypothesis}\label{sec:B-3}

Two of the Levi relations are positively associated with whole-compound transparency in Bell and Schäfer's models: IN and FOR. Drawing on Bell and Plag (2012), Bell and Schäfer suggest a possible mechanism: if more transparent constituents are less informative, and if compound stress falls on the more informative constituent, the IN/FOR effects on transparency would line up with the stress patterns observed for those compounds \autocite[168]{bell-schafer-2016-modelling-transparency}. The connection is offered as a hypothesis that further work would need to test, not as an established mechanism.

\subsection{\textup{N1}/\textup{N2} asymmetry}\label{sec:B-4}

Although whole-compound transparency is a function of the transparencies of the two constituents, \textup{N1} (the modifier) and \textup{N2} (the head) differ in the nature of their contribution. The asymmetry is one of the two headline conclusions of Bell and Schäfer's paper. The other is that all the significant predictors of rated transparency in their models are also independent predictors of processing speed, leading them to a tentative hypothesis: \enquote{perceived transparency may itself be a reflex of ease of processing} \autocite[157, abstract]{bell-schafer-2016-modelling-transparency}. The hypothesis is offered as a direction for further work, not a result of this study.

\subsection{The Reddy et al.~database}\label{sec:B-5}

The empirical anchor of Bell and Schäfer's models is the database of Reddy and colleagues (2011), used as the dependent variable. The database consists of 30 transparency ratings for each of 90 two-part English NN compounds, totalling 8,100 ratings, collected via Amazon Mechanical Turk. Annotators rated, on a 0--5 scale, how literal each constituent was in the compound and how literal the compound as a whole was. Ratings are continuous; they don't partition compounds into transparent-versus-opaque classes \autocite[164]{bell-schafer-2016-modelling-transparency}.
